Speed and strength of an epidemic intervention

### 2020-02-02 doc/speedstrength.tex
comments

An epidemic can be described by its \emph{speed} and \emph{strength}.
The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
Epidemic is characterized by the reproductive number \RR, which measures how many new cases are caused by a typical \emph{individual} case.
Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.

@@

\jd{I'm horrified that you would suggest saying ``exponentially'' in a non-technical sense in a scientific paper. Should we think about showing one or three generation intervals in panels a and c (intrinsic and backward respectively) to help people think about averaging?}

@@

Using the speed-based decomposition provides a valuable perspective on this situation.
The hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but peaks very fast (because sexually active individuals are the most likely to seek testing); 
the assumed hazard rate goes down after a few months to reflect the satuaration assumptions discussed above. 
The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
The effective speed of the intervention stays relatively constant, because \jd{WHY? Can we understand this part better? I was going to say it is driven by the hazard of early detection, but that hazard is higher than I expected compared to the speed.}  It increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.

@@

The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive below 1.
In this study, we extend this idea by first defining the strength of an intervention on the same scale as the strength of an epidemic (i.e., basic reproductive number $\mathcal R_0$); and then constructing a parallel interpretation which measures the speed of an intervention on the same scale as the speed of an epidemic (i.e., exponential growth rate $r$).
We show that an intervention can successfully contain an epidemic if either its strength or speed is greater than their corresponding measures of the epidemic.

We compare two HIV intervention strategies (condom intervention and test-and-treat intervention) to demonstrate that both strength- and speed-based frameworks can be useful in evaluating the effectiveness of an epidemic intervention.
In particular, we recapitulate the result of \cite{eaton2014proportion} who used detailed mathematical modeling of HIV transmission to show that the amount of early transmission does not affect the effectiveness of the ART.
While they gave a strength-based argument for the result --  increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion} -- there is a simple speed-based intuition behind this phenomenon: in principle, we can control the outbreak if we can identify infected individuals and put them on ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the amount of early transmission.
While both strength- and speed-based frameworks give the same conclusion about the outcome of an intervention, one can provide a clearer interpretation of the outcome than the other, depending on the situation.

>>>>

An epidemic can be described by its \emph{speed} and \emph{strength}.
The speed of an epidemic is characterized by the exponential growth rate $r$, which measures how \emph{fast} an epidemic grows at the population level.
The strength of an epidemic is characterized by the reproductive number \RR, which measures how many new cases are caused by a typical \emph{individual} case.
Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.

@@

\jd{I'm horrified that you would suggest saying ``exponentially'' in a non-technical sense in a scientific paper. Should we think about showing one or three generation intervals in panels a and c (intrinsic and backward respectively) to help people think about averaging?}
\swp{I meant to say approximately exponentially. Based on panel a, it does look like it grows approximately linearly on a log-scale and this rate approximately matches the maximum hazard rate.}

@@

Using the speed-based decomposition provides a valuable perspective on this situation.
The hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but peaks very fast (because sexually active individuals are the most likely to seek testing); 
the assumed hazard rate goes down after a few months to reflect the satuaration assumptions discussed above. 
The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
The effective speed of the intervention stays relatively constant, because \jd{WHY? Can we understand this part better? I was going to say it is driven by the hazard of early detection, but that hazard is higher than I expected compared to the speed.}  It increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
\swp{I have an explanation. The backward GI distribution is incredibly narrow. Its mean is 0.97 years but the distribution peaks around 0.2 years. Therefore, the inital superexponentiality actually matters quite a lot when we try to average it out, and therefore we get something that's much lower than the peak hazard. If we made the initial $h$ constant, rather than starting it from 0, the speed of intervention would match the hazard of early detection much better.}

@@

The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number below 1.
In this study, we extend this idea by first defining the strength of an intervention on the same scale as the strength of an epidemic (i.e., basic reproductive number $\mathcal R_0$); and then constructing a parallel interpretation which measures the speed of an intervention on the same scale as the speed of an epidemic (i.e., exponential growth rate $r$).
We show that an intervention can successfully contain an epidemic if either its strength or speed is greater than their corresponding measures of the epidemic.

We compare two HIV intervention strategies (condom intervention and test-and-treat intervention) to demonstrate that both strength- and speed-based frameworks can be useful in evaluating the effectiveness of an epidemic intervention.
In particular, we recapitulate the result of \cite{eaton2014proportion} who used detailed mathematical modeling of HIV transmission to show that the amount of early transmission does not affect the effectiveness of the ART.
While they gave a strength-based argument for the result --  increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion} -- there is a simple speed-based intuition behind this phenomenon: in principle, we can control an outbreak if we can identify infected individuals and treat them with ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the amount of early transmission.
While both strength- and speed-based frameworks give the same conclusion about the outcome of an intervention, one can provide a clearer interpretation of the outcome than the other, depending on the situation.
